\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-16 doc/mlsys_paper/paper.tex
A first attempt at refactoring the templates section.
Experiments are not yet done.

\documentclass{article}
\usepackage[preprint]{nips_2018}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage[newfloat]{minted}
\usepackage{caption}
\usepackage{parcolumns}
\usepackage{booktabs}

@@

\vspace*{-0.3em}
\section{Templates for speed}
\vspace*{-0.5em}

Many optimization toolkits depend on the user passing in a function pointer to a
function that needs to be optimized.  For instance, the {\tt scipy.optimize}
package in Python can be used like this:

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{python}
    def rosen(x): return 100.0 * (x[2] - x[1]**2.0)**2.0 + (1 - x[1])**2.0)
    res = scipy.optimize.minimize(rosen, [-1.2, 1], method='nelder-mead')
\end{minted}
\vspace*{-0.3em}

During the call to {\tt minimize()}, the {\tt rosen()} function will be called
repeatedly---but it will be called as a function pointer, which must be
dereferenced each time {\tt rosen()} is called.  In addition, some
optimizations such as inlining (and many optimizations enabled by inlining) are
not possible when using function pointers.  Since some optimization algorithms
need to compute the objective very many times, the overhead of the lookup and
missed optimization opportunities may be non-negligible.  This applies even for
SGD-like optimizers that train machine learning models on very large datasets!

Since {\tt ensmallen} is written in C++ with templates, the function pointer
dereferencing overhead can be avoided and the compiler can choose to inline the
function call and perform additional optimizations when possible, such as
temporary variable elimination and dead code pruning.

\vspace*{-0.3em}
\section{Automatic Generation of Methods}
\vspace*{-0.5em}

@@

% This is up here so that it's on the right page...
\begin{table}[t]
\begin{center}
\begin{tabular}{cccc}
\toprule
{\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
{\bf 0.004s} & 1.044s & 0.021s & 2.920s \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with different
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.}
\label{tab:rosenbrock_results}
\end{table}

@@

Since all of this code generation is automatically done at compile-time and not
at runtime, the compiler is able to perform dead code elimination
on the unused result, and may be able to avoid unnecessary calculations.  In some
cases, the generated {\tt Evaluate()} or {\tt Gradient()} functions can be
as fast as their corresponding hand-written versions.

%% CS: a subset of the text from the problematic "Templates for speed" section above
%% CS: can be placed here

Overall, the automatic code generation functionality in {\tt ensmallen}
reduces the requirements for users
when they are implementing their own objective functions to be optimized.
This leads to both quicker development and reduces the likelihood of bugs.

@@

First, we consider the added overhead of other frameworks.  For our experiment,
we choose the simple and popular Rosenbrock function~\cite{Rosenbrock1960}:
$f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2)$.  For the optimizer, we use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer.  Simulated annealing will call the objective function
very many times; for each simulation we limit the optimizer to 100k
objective evaluations.  We compare four frameworks for this task:

@@

Table~\ref{tab:rosenbrock_results} shows the average runtime
across 10 trials for $100000$ iterations of simulated annealing with each
implementation.  The simulations were run on a % TODO: what are Marcus's
% system specs?
{\tt \$COMPUTER}.
Of these frameworks, only Julia and {\tt ensmallen} are able to avoid the
function pointer dereference and take advantage of inlining and related
optimizations---and this difference is strongly reflected in the results.

>>>>

\documentclass{article}
\usepackage[preprint]{nips_2018}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage[newfloat]{minted}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{parcolumns}
\usepackage{booktabs}

@@

\vspace*{-0.3em}
\section{Template Metaprogramming for Ease of Use and Efficiency}
\vspace*{-0.5em}

@@

% This is up here so that it's on the right page...
\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & {\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
default & {\bf 0.004s} & 1.044s & 0.021s & 2.920s \\
tuned & & 0.5s-ish? & & ?? \\ % TODO
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with different
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.  The {\it tuned} row indicates that the code was manually
modified for speed.}
\label{tab:rosenbrock_results}
\end{table}

@@

The use of templates in this manner also allows for compiler optimizations that
would not otherwise be possible (and that are often not possible in other
frameworks).  Firstly, because the objective function class itself is a template
parameter, the compiler is able to avoid the overhead of a function pointer
dereference, which would not be easily possible when using a language with
virtual inheritance.  The compiler is also able to use inlining and any
optimizations that may imply, including removing temporary value and dead code
elimination.  Further, if {\tt ensmallen} automatically generates an {\tt
Evaluate()} or {\tt Gradient()} method from a user-supplied {\tt
EvaluateWithGradient()} method, the compiler can in some cases recognize and
remove the computation of unnecessary results.  For instance, in an
automatically generated {\tt Evaluate()} method, the computation of the gradient
from {\tt EvaluateWithGradient()} can be avoided entirely.

Overall, the automatic code generation functionality in {\tt ensmallen}
reduces the requirements for users
when they are implementing their own objective functions to be optimized,
and allows users a way to provide more efficient implementations of their
objective functions.
This leads to quicker development, quicker results, and reduces the likelihood of bugs.

@@

For our first experiment, we consider the simple and popular Rosenbrock
function~\cite{Rosenbrock1960}: $f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 -
x_1^2)$.  For the optimizer, we use simulated
annealing~\cite{kirkpatrick1983optimization}, a gradient-free optimizer.
Simulated annealing will call the objective function very many times; for each
simulation we limit the optimizer to 100k objective evaluations.  Since the
objective function is so simple and is called so many times, this can help us
understand the overheads of different frameworks.

In our comparisons we use four different frameworks:

@@

% TODO: get Marcus' system specs.
We implemented and ran our code on a {\tt \$COMPUTER} running an {\tt
\$OPERATING\_SYSTEM} with gcc version {\tt \$GCC\_VERSION}, Julia version {\tt
\$JULIA\_VERSION}, Python {\tt \$PYTHON\_VERSION}, and Octave {\tt
\$OCTAVE\_VERSION}.

Initially, we implemented these functions as simply as possible and ran them
without any tuning, which reflects how a normal user might interact with the
software.  These results are seen in the first row of
Table~\ref{tab:rosenbrock_results}.  Of the four frameworks, only Julia and {\tt
ensmallen} are compiled, and thus are able to avoid the function pointer
dereference and take advantage of inlining and related optimizations.

However, both Python and GNU Octave have routes for acceleration, such as
numba~\cite{TODO} and Octave's built-in JIT.  Therefore, we hand-optimized the
Rosenbrock implementation using numba and ran the Octave implementation with the
JIT.  For the numba implementation, the underlying {\tt anneal.anneal()}
function had to be significantly modified.  These did produce some acceleration,
as can be seen in the second row of Table~\ref{tab:rosenbrock_results}.
